```{r setup, eval=TRUE, echo=FALSE}
library(knitr)
#opts_knit$set(upload.fun = image_uri)
opts_knit$set(upload.fun = imgur_upload)


```
[![Build Status](https://travis-ci.org/fmichonneau/rotl.svg)](https://travis-ci.org/fmichonneau/rotl)
[![Build status](https://ci.appveyor.com/api/projects/status/57rapgc0s7fwxwad?svg=true)](https://ci.appveyor.com/project/fmichonneau/rotl)
[![Coverage Status](https://coveralls.io/repos/fmichonneau/rotl/badge.svg)](https://coveralls.io/r/fmichonneau/rotl)

# An R interface to Open Tree API

`rotl` is an R package to interact with the Open Tree of Life data APIs. It was
initially developed as part of the
[NESCENT/OpenTree/Arbor hackathon](http://blog.opentreeoflife.org/2014/06/11/apply-for-tree-for-all-a-hackathon-to-access-opentree-resources/).

Client libraries to interact with the Open Tree of Life API also exists for
[Python](https://github.com/OpenTreeOfLife/opentree-interfaces/tree/master/python)
and [Ruby](https://github.com/SpeciesFileGroup/bark).


## Installation

If you want to install this package, you first need to install
[devtools](https://github.com/hadley/devtools), as `rotl` is not yet available
on CRAN.

`rotl` uses [rncl](https://github.com/fmichonneau/rncl) to parse trees, so you
will also need to install that package.

```{r, eval=FALSE}
install.packages("devtools")
```

Then you can install `rotl` using:

```{r, eval=FALSE}
library(devtools)
install_github("fmichonneau/rncl")
install_github("fmichonneau/rotl", dependencies = TRUE, build_vignette=TRUE)
```

To build the vignette, you will need to have pandoc 1.12.3 or higher
installed. If you use
[RStudio](https://www.rstudio.com/products/rstudio/download/), make sure you
have v0.99 (as it comes with pandoc 1.13.1); if you don't use RStudio, you can
find pandoc for your operating system [here](http://pandoc.org/installing.html)

## Vignettes

There are two vignettes:

- start by checking out the "How to use `rotl`?" by typing:
  `vignette("how-to-use-rotl", package="rotl")` after installing the
  package.

- then explore how you can use `rotl` with other packages to combine your data
  with trees from the Open Tree of Life project by typing:
  `vignette("data_mashups",package="rotl")`.

## Quick start

### Get a little bit of the big Open Tree tree

Taxonomic names are represented in the Open Tree by numeric identifiers, the
`ott_ids` (Open Tree Taxonomy identifiers). To extract a portion of a tree from
the Open Tree, you first need to find `ott_ids` for a set of names using the
`tnrs_match_names` function:

```{r resolve}
library(rotl)
apes <- c("Pan", "Pongo", "Pan", "Gorilla", "Hylobates", "Hoolock", "Homo")
(resolved_names <- tnrs_match_names(apes))
```

Now get the tree with just those tips:

```{r get_tr}
tr <- tol_induced_subtree(ott_ids=resolved_names$ott_id)
plot(tr)
```

### Code of Conduct

Please note that this project is released with a
[Contributor Code of Conduct](CONDUCT.md). By participating in this project you
agree to abide by its terms.
