```{r setup, eval=TRUE, echo=FALSE}
library(knitr)
#opts_knit$set(upload.fun = image_uri)
opts_knit$set(upload.fun = imgur_upload)


```
[![Build Status](https://travis-ci.org/fmichonneau/rotl.svg)](https://travis-ci.org/fmichonneau/rotl)
[![Coverage Status](https://coveralls.io/repos/fmichonneau/rotl/badge.svg)](https://coveralls.io/r/fmichonneau/rotl)

# An R interface to Open Tree API

`rotl` is an R package to interact with the Open Tree of Life data APIs. It was
initially developed as part of the
[NESCENT/OpenTree/Arbor hackathon](http://blog.opentreeoflife.org/2014/06/11/apply-for-tree-for-all-a-hackathon-to-access-opentree-resources/).

Client libraries to interact with the Open Tree of Life API also exists for
[Python](https://github.com/OpenTreeOfLife/opentree-interfaces/tree/master/python)
and [Ruby](https://github.com/SpeciesFileGroup/bark).


## Installation

If you want to install this package, you first need to install
[devtools](https://github.com/hadley/devtools), as `rotl` is not yet available
on CRAN.

`rotl` uses [rncl](https://github.com/fmichonneau/rncl) to parse trees, so you
will also need to install that package.

```{r, eval=FALSE}
install.packages(c("devtools", "rncl"))
```

Then you can intall `rotl` using:

```{r, eval=FALSE}
library(devtools)
install_github("fmichonneau/rotl")
```

## Vignettes

There are two vignettes:
- start by checking out the "How to use `rotl`?" by typing:
  `vignette("how-to-use-rotl", package="rotl")` after installing the package.
- explore in a little more detail how to use `rotl` with other packages with
  [this vignette](http://dwinter.github.io/rotl-vignette/)

## Simple examples

Note: the library is still in active development and behaviour of the following
functions may well change in the future:

### Get a little bit of the big Open Tree tree

First find ott ids for a set of names:

```{r resolve}
library(rotl)
apes <- c("Pan", "Pongo", "Pan", "Gorilla", "Hylobates", "Hoolock", "Homo")
(resolved_names <- tnrs_match_names(apes))
```
Now get open tree to return a tree with just those tips.


```{r get_tr}
tr <- tol_induced_subtree(ott_ids=resolved_names$ott_id)
plot(tr)
```


### Find trees focused on my favourite taxa

```{r mammals}
furry_studies <- studies_find_studies(property="ot:focalCladeOTTTaxonName", value="Mammalia")
( furry_ids <- unlist(furry_studies$matched_studies) )

```

### Get a specific study tree
```{r tree}
library(ape)
furry_metadata <-get_study_meta(2647)
furry_metadata$nexml$treesById
furry_tr <- get_study_tree(study_id="2647", tree_id="tree6169")
plot(furry_tr)
```

### Code of Conduct

Please note that this project is released with a
[Contributor Code of Conduct](CONDUCT.md). By participating in this project you
agree to abide by its terms.
