---
title: "Connecting data to Open Tree trees"
author: "David Winter"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connecting data to Open Tree trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Combining data from OToL and other sources.

One of the major goals of `rotl` is to help users combine data from other
sources with the phylogenetic trees in the Open Tree database. This examples
demonstrates how a user might connect data they have collected to trees from
Open Tree.

## Get Open Tree ids to match your data.

Let's say you have a dataset where each row represents a measurement taken from
one species, and your goal is to put these measurements in some phylogenetic
context. Here's a small example: the best estimate of the mutation rate for a
set of unicellular Eukaryotes along with some other property of those species
which might explain the mutation rate:

```{r, data}
csv_path <- system.file("extdata", "protist_mutation_rates.csv", package = "rotl")
mu <- read.csv(csv_path, stringsAsFactors=FALSE)
mu
```

If we want to get a tree for these species we need to start by finding the
unique ID for each of these species in the Open Tree database. We can use the
Taxonomic Name Resolution Service (`tnrs`) functions to do this. Before we do
that we should see if any of the taxonomic contexts, which can be used to narrow
a search and avoid conflicts between different codes, apply to our group of species:

```{r, context}
library(rotl)
tnrs_contexts()
```

Hmm, none of those groups contain all of our species. In this case we can
search using the `All life` context and the function `tnrs_match_names`:

```{r, match}
taxon_search <- tnrs_match_names(names=mu$species, context_name="All life")
knitr::kable(taxon_search)
```

Good, all of our  species are known to Open Tree. Note, though, that one of the names
is a synonym. _Saccharomyces pombe_ is older name for what is now called
_Schizosaccharomyces pombe_. As the name suggests, the Taxonomic Name
Resolution Service is designed to deal with these problems (and similar ones
like misspellings), but it is always a good idea to check the results of
`tnrs_match_names` closely to ensure the results are what you expect.

In this case we have a good ID for each of our species so we can move on. Before
we do that, let's ensure we can match up our original data to the Open Tree
names and IDs by adding them to our `data.frame`:

```{r, munge}
mu$ott_name <- taxon_search$unique_name
mu$ott_id <- taxon_search$ott_id
```

## Find a tree with your taxa

Now let's find a tree. There are two possible options here: we can search for
published studies that include our taxa or we can use the 'synthetic tree' from
Open Tree. We can try both approaches.

### Published trees

Before we can search for published studies or trees, we should check out the
list of properties we can use to perform such searches:

```{r, properties}
studies_properties()
```

We have `ottIds` for our taxa, so let's use those IDs to search for trees that
contain them.  Starting with our first species _Tetrahymena thermophila_ we can
use `studies_find_trees` to do this search.

```{r taxon_count}
studies_find_trees(property="ot:ottId", value="180195")
```
Well... that's not very promising. We can repeat that process for all of the IDs
to see if the other species are better represented.


```{r, all_taxa_count}
hits <- lapply(mu$ott_id, studies_find_trees, property="ot:ottId", detailed = FALSE)
sapply(hits, function(x) sum(x[["n_matched_trees"]]))
```
OK, most of our species are not in any of the published trees available. You can
help fix this sort of problem by [making sure you submit your published trees to
Open Tree](https://tree.opentreeoflife.org/curator).

### A part of the synthesis tree

Thankfully, we can still use the complete Tree of Life made from the
combined results of all of the published trees and taxonomies that go into Open
Tree. The function `tol_induced_subtree` will fetch a tree relating a set of IDs.

Using the default arguments you can get a tree object into your R session:


```{r subtree,  fig.width=7, fig.height=4}
tr <- tol_induced_subtree(ott_ids=mu$ott_id)
plot(tr)
```

### Connect your data to the tips of your tree

Now we have a tree for of our species, how can we use the tree and the data
together?

The package `phylobase` provide an object class called `phylo4d`, which is
designed to represent a phylogeny and data associated with its tips. In oder to
get our tree and data into one of these objects we have to make sure the labels
in the tree and in our data match exactly. That's not quite the case at the
moment (tree labels have underscores and IDs appended):

```{r, match_names}
mu$ott_name[1]
tr$tip.label[4]
```

We can use `sub` to the remove the underscores and `ottId` from the tree
(check out `?regex` to see how these patterns work) and `%in%` to confirm that
each of the modified labels matches a taxon in our `data.frame`:

```{r, sub}
tr$tip.label <- sub("_ott\\d+", "", tr$tip.label)
tr$tip.label <- sub("_", " ", tr$tip.label)
tr$tip.label %in% mu$ott_name
```

Ok, now the tips are together we can make a new dataset. The `phylo4d()`
functions matches tip lables to the row names of a `data.frame`, so let's make
a new dataset that contains just the relevant data and has row names to match
the tree



```{r phylobase}
library(phylobase)
mu_numeric <- mu[,c("mu", "pop.size", "genome.size")]
rownames(mu_numeric) <- mu$ott_name
tree_data <- phylo4d(tr, mu_numeric)
```
And now we can plot the data and the tree together


```{r,  fig.width=7, fig.height=5}
plot(tree_data)
```

## What next

The demonstration get's you to the point of visualizing your data in a
phylogenetic context. But there's a lot more you do with this sort of data in R.
For instance, you could use packages like `ape`, `caper`, `phytools` and
`mcmcGLMM` to perform phylogenetic comparative analyses of your data. You could
gather more data on your species using packages that connect to
trait databases like `rfishbase`, `AntWeb` or `rnpn` which provides data from
the US National Phenology Network. You could also use `rentrez` to find genetic
data for each of your species, and use that data to generate branch lengths for
the phylogeny.
