---
title: "How to use rotl?"
author: "François Michonneau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`rotl` provides an interface to the Open Tree of Life (OTL) API and allows users
to query the API, retrieve parts of the Tree of Life and integrate these parts
with other R packages.

The OTL API provides service to access:

* the **Tree of Life** a.k.a. TOL (the synthetic tree): a single draft tree that is
a combination of **the OTL taxonomy** and the **source trees** (studies)
* the **Graph of Life** a.k.a. GOL: the neo4j graph database used to store and
build the synthetic tree. It includes additional information (more nodes and
edges) that is not in the TOL.
* the **Taxonomic name resolution services** a.k.a. TNRS: the methods for
  resolving taxonomic names to the internal identifiers used by the TOL and the
  GOL (the `ott ids`).
* the **Taxonomy** a.k.a. OTT (for Open Tree Taxonomy): which represents the
  synthesis of the different taxonomies used as a backbone of the TOL when no
  studies are available.
* the **Studies** containing the source trees used to build the TOL, and
extracted from the scientific literature.

In `rotl`, each of these services correspond to functions with different
prefixes:

| Service       | `rotl` prefix |
|---------------|---------------|
| Tree of Life  | `tol_`        |
| Graph of Life | `gol_`        |
| TNRS          | `tnrs_`       |
| Studies       | `studies_`    |



## Demonstration of a basic workflow

The most common use for `rotl` is probably to start from a list of species and
get the relevant parts of the tree for these species. This is a two step
process:

1. the species names need to be matched to their `ott_id` (the Open Tree
Taxonomy identifiers) using the Taxonomic name resolution services (TNRS)
1. these `ott_id` will then be used to retrieve the relevant parts of the Tree
of Life.

### Step 1: Matching taxonomy to the `ott_id`

Let's start by doing a search on a diverse group of taxa: a tree frog (genus
_Hyla_), a fish (genus _Salmo_), a sea urchin (genus _Diadema_), a nautilus
(genus _Nautilus_), and an oak tree (genus _Acer_).

```{r}
library(rotl)
taxa <- c("Hyla", "Salmo", "Diadema", "Nautilus", "Acer")
resolved_names <- tnrs_match_names(taxa)
```

It's always a good idea to check that the resolved names match what you
intend:

`r knitr::kable(resolved_names)`

The column `unique_name` sometimes indicates the higher taxonomic level
associated with the name. Here we queried on genera names, so the API indicates
the family names associated with some the genera. The column `number_matches`
indicates the number of `ott_id` that corresponds to a given name. In our
example, our search on _Diadema_ returns 4 matches, and the one returned by
default is a fungus and not the sea urchin that we want for our query.

The argument `context_name` allows you to limit the taxonomic scope of your
search. In this case, if we limit our search to animals, the genus _Diadema_ now
matches our sea urchin:

```{r}
resolved_names <- tnrs_match_names(taxa, context_name = "Animals")
```

`r knitr::ktable(resolved_names)`

If you are trying to build a tree with deeply divergent taxa and run into this
issue, see "How to change the ott ids assigned to my taxa?" in the FAQ below.


### Getting the tree corresponding to our taxa

Now that we have the correct `ott_id` for our taxa, we can ask for the tree. By
default, the object returned by `tol_induced_subtree` is a phylo object (from
the [ape](http://cran.r-project.org/package=ape) package), so we can plot it
directly.

```{r, fig.width=7, fig.height=4}
my_tree <- tol_induced_subtree(ott_ids = resolved_names$ott_id)
plot(my_tree, no.margin=TRUE)
```


## FAQ

### How to change the ott ids assigned to my taxa?

If you realize that `tnrs_match_names` assigns the incorrect taxonomic group to
your name (e.g., because of synonymy) and changing the `context_name` does not
help, you can use the function `inspect_match_names`. This function takes the
object resulting from using `tnrs_match_names`, and either the row number, the
taxon name (you used in your search in lowercase), or the `ott_id` returned by
the initial query.

To illustrate this, let's re-use the previous query but by adding a fungus to
our list of taxa, such that restricting our search to animals doesn't work.

```{r}
taxa <- c("Hyla", "Salmo", "Diadema", "Nautilus", "Acer", "Boletus")
resolved_names <- tnrs_match_names(taxa)
inspect_match_names(resolved_names, taxon_name = "diadema")
```

In our case, we want the second row in this data frame to replace the
information that initially matched for _Diadema_. We can now use the
`update_match_names`, to change to the correct taxa (the sea urchin not the
fungus):

```{r}
resolved_names <- update_match_names(resolved_names, taxon_name = "diadema",
                                     new_row_number = 2)

## we could also have used the ott_id to replace this taxon:
## resolved_names <- update_match_names(resolved_names, taxon_name = "diadema",
##                                     new_ott_id = 631176)
```

And now our `resolved_names` data frame includes the taxon we want:

`r knitr::kable(resolved_names)`

### How do I know that the taxa I'm asking for is the correct one?

### How do I find trees focused on my favourite taxa?

### How do I get a specific tree from a study?
